---
title: "Assignment3A-KCH-Approach"
author: "Kailot C. Harris"
date: "2026/09/15"
format: html
theme: cosmo
editor: visual
---

# Introduction and Overview

This assignment serves as an extension of Assignment 2A. Here, the pre-existing data pipeline for survey response ingestion into a database is already in place. The important objective for this assignment is determining a global baseline estimate for movies with missing ratings. Some important steps along the way include importing the data from a PostgreSQL database, tidying the data, and calculating necessary parameters.

Anticipated challenges in this assingment include developing concise R implementations to perform calculations, vector manipulation and finding missing ratings in the database to as placeholders to generate predictions.

# Codebase

## Database Connection

```{r connect-db}
#| echo: true
#| message: false
#| warning: false

library(RPostgres)

#establish connection
con <- dbConnect(RPostgres::Postgres(),
                 dbname = 'Second_DB_DATA607', 
                 host = 'localhost',
                 port = 5432,
                 user = Sys.getenv("DATA_607_DB_USER"),
                 password = Sys.getenv("DATA_607_DB_PW"))
```

## Data Import and Database Disconnection

```{r import-data-from-db}
#| echo: TRUE
#| warning: FALSE
#| message: FALSE


library(tidyverse)

df_movies <- dbGetQuery(con, "select * from movies;")

df_respondents <- dbGetQuery(con, "select * from respondents;")

df_ratings <- dbGetQuery(con, "select * from ratings;")

##closing the DB conneciton after importing data
dbDisconnect(con)
## Confirming df_Ratings is of expected dimensions 
glimpse(df_ratings)
```

## Data Tidying

```{r join-tables}

df <- left_join(df_ratings, 
                df_movies, 
                by = "movie_id")

df <- left_join(df, 
                df_respondents, 
                by = "respondent_id")

head(df)
```

## Missing Data

```{r missing-data}

respondent_movie_combinations <- expand.grid(
  respondent_id = df_respondents$respondent_id,
  movie_id = df_movies$movie_id
)  

df_2 <- respondent_movie_combinations |>
  left_join(df_ratings, 
            by = c("respondent_id",
                   "movie_id")) |>
  left_join(df_movies, 
            by="movie_id")

#NA values visible in rating id and rating column
head(df_2)

```

## Global Mean Determination

```{r global-mean-calcs}

mu <- df |>
  group_by(movie_id) |>
  summarize (mean_rating_movie = mean(rating, na.rm=TRUE)) |>
  summarize (global_mean = mean(mean_rating_movie))

mu[1,1]
```

```{r user-contributions}
b_user <- df |>
  group_by(respondent_id) |>
  summarize (mean_rating_respondent = mean(rating, na.RM=TRUE)) |>
  mutate(b_user = mean_rating_respondent - as.numeric(mu[1,1]))

b_user <- b_user |>
  select(respondent_id, b_user)

b_user
```

```{r movie-contributions}
b_movie <- df |>
  group_by(movie_id) |>
  summarize(b_movie = mean(rating, na.RM = TRUE)-as.numeric(mu[1,1]))

b_movie
```

## Estimating Movie Rating for Users

### Finding missing ratings

```{r}
df_missing <- df_2 |>
  filter(is.na(rating))

head(df_missing)
```

It seems that respondents 4, 6, and 7 haven't seen "One Battle After Another (2025)" yet. Global baseline estimates will be generated for all mock respondents for any movie they left unrated.

```{r estimate-generation}

mu <- as.numeric(mu[1,1])
  
df_estimates <- df_missing |>
  left_join(b_user, by = "respondent_id") |>
  left_join(b_movie, by = "movie_id") |>
  mutate(global_mean = mu,
         rating_est = global_mean+b_user+b_movie)

df_estimates |> select(respondent_id, movie_id, title, rating_est)

df_estimates
```
